A thermodynamic view of dusty protoplanetary disks

### 2017-05-23 linear.tex
wording edits in SI section (main text and new analytic model in appendix)

\subsection{Streaming instability}\label{si}
We now consider 3D but non-self-gravitating disks.  
In the special case of an unstratified disk (neglecting the vertical
component of the stellar gravity), we may also Fourier analyze in $z$
to obtain an algebraic dispersion relation of the form  
$\sum_{j=0}^{5}c_j(k_x,k_z)\sigma^j = 0$, where $k_z$ is a real
vertical wavenumber. The coefficients $c_j$ can be read 
off Eq. \ref{streaming_dispersion} in  Appendix \ref{compressible_streaming}. 
There we also show that this dispersion relation reduces to that for
the streaming instability (SI) in the limit of incompressible gas and small
$\tstop$ \citep{youdin05a,jacquet11}.

@@

We solve the full dispersion relation, Eq. \ref{streaming_dispersion} for selected cases where SI  
growth rates have been 
calculated analytically from full two-fluid disk models and confirmed
with particle-gas numerical simulations 
\citep[namely][]{youdin07b,bai10b}. These authors use normalized
wavenumbers $K_{x,z} = \eta r k_{x,z}$ where
\begin{align} 
  \eta \equiv -\frac{1}{2\rhog r\OmK^2}\frac{\p P}{\p r} = 
  \frac{1}{2\left(1-\tepsilon\right)}\frac{F_r}{r\OmK^2}, 
\end{align} 
measures the pressure offset of Keplerian rotation. We use
$\eta=0.05c_s/r\OmK$. They also define the stopping time as
$\tau_\mathrm{s}=\tstop/(1-\tepsilon)$.

Table \ref{si_compare} compares the eigenfrequencies obtained from the
one-fluid dispersion relation to that from the above studies. As  
expected eigenfrequencies agree better with decreasing $\tstop$ since
in that limit the mixture behaves more like a single fluid. Most 
importantly, we find the work done $\mathcal{W}>0$ in all cases, and
hence find growing oscillations.

@@

We also checked the growth rates $s$ satisfy 
\begin{align} 
  s = \frac{\left|\sigma\right|^2\imag\left(\Delta P
    \Delta\rho^*/\rho\right)}{2\real\left(\sigma\right)\rho\left(\left|\dd
  v_x\right|^2+\left|\dd
  v_z\right|^2\right)}, \label{si_check}
\end{align}
as implied by Eq. \ref{thermal_instability} and
Eq. \ref{pdv}. %, or directly from
%Eq. \ref{streaming_mass}---\ref{streaming_vz}. 
Interestingly, we find in dust-rich disks with 
$\epsilon>1$, Eq. \ref{si_check} can be satisified with $\Delta
P\simeq \ii\dd v_x \p_rP/\sigma$, i.e. the radial pressure gradient
is responsible for growth. Conversely, for `linB' with $\epsilon
< 1$, one can approximate $\Delta P \simeq \dd P$ in Eq. \ref{si_check}. %This
%suggests that for SI in gas-dominated disks, the distinction between
%Eulerian and Lagragian pressure perturbations is unimportant. 
%{\bf but weak growth in this case}
%However
%in that case the gro

In Table \ref{si_compare} we also calculate the phase difference
between the Lagragian pressure perturbation and density perturbations
as    
\begin{align*} 
\varphi \equiv -\sgn(\omega)\arg\left(\Delta P\Delta\rho^*\right)
        = \sgn(\omega)\arg\left(\Delta\rhog^*\Delta\rhod\right),\notag
\end{align*}
%where we have applied the isothermal equation of state for the second equality.
then $\varphi > 0 $ indicates gas density lagging behind dust density,
which is true for all the cases. In particular, `linB' indicates that
growth rates vanish as $\varphi \to 0$.  Thus the streaming
instability is indeed associated with a phase lag between gas and dust
density evolutions.

In a future work we will perform a detailed comparison between the
simplified one-fluid and full two-fluid frameworks in calculating SI.

\begin{figure}
\includegraphics[width=linewidth]{figures/si_2f_1f_compare}
\caption{{\bf in case the paper isn't long enough} Comparison of the linear
streaming instability obtained from a full two-fluid analysis and the current 
one-fluid framework simplified by the terminal velocity approximation. The vertical
wavenumber is fixed and growth rates are maximized over $K_x$. 
\label{si_compare}}
\end{figure}
%"generalized stokes number" would include other problem parameters,
%e.g. wavenumber

@@

>>>>

\subsection{Streaming instability}\label{si}
We now consider 3D disks without self-gravity. We neglect the vertical     
component of the stellar gravity, appropriate for studying regions
near the disk midplane. This  
allows us to Fourier analyze in $z$ to obtain 
an algebraic dispersion relation of the form  
$\sum_{j=0}^{5}c_j(k_x,k_z)\sigma^j = 0$, where $k_z$ is a real
vertical wavenumber. The coefficients $c_j$ can be read 
off Eq. \ref{streaming_dispersion} in  Appendix \ref{compressible_streaming}. 
There we also show that this dispersion relation reduces to that for
the streaming instability (SI) in the limit of incompressible gas and small
$\tstop$ \citep{youdin05a,jacquet11}.

@@

In Table \ref{si_compare} we solve the full dispersion relation
(Eq. \ref{streaming_dispersion}) numerically for selected cases where
analytic SI growth rates have been verified with particle-gas
numerical simulations \citep[namely][]{youdin07b,bai10b}.  
Following previous works on SI, we use normalized 
wavenumbers $K_{x,z} = \eta r k_{x,z}$ where
\begin{align} 
  \eta \equiv -\frac{1}{2\rhog r\OmK^2}\frac{\p P}{\p r} = 
  \frac{1}{2\left(1-\tepsilon\right)}\frac{F_r}{r\OmK^2}, 
\end{align} 
measures the pressure offset of Keplerian rotation. We fix 
$\eta=0.05c_s/r\OmK$. In this section section we also quote the
particle stopping time $\tau_\mathrm{s}=\tstop/(1-\tepsilon)$.

The eigenfrequencies obtained from the one-fluid dispersion relation 
are compared that from a full, two-fluid analysis \citep[similar 
to][]{youdin05a, kowalik13}. As expected eigenfrequencies agree better
with decreasing $\taus$ since in that limit the mixture behaves more 
like a single fluid. Most importantly, we find the work done
$\mathcal{W}>0$ in all cases, and hence find growing oscillations.

@@

% We also checked the growth rates $s$ satisfy 
% \begin{align} 
%   s = \frac{\left|\sigma\right|^2\imag\left(\Delta P
%     \Delta\rho^*/\rho\right)}{2\real\left(\sigma\right)\rho\left(\left|\dd
%   v_x\right|^2+\left|\dd
%   v_z\right|^2\right)}, \label{si_check}
% \end{align}
% as implied by Eq. \ref{thermal_instability} and
% Eq. \ref{pdv}. %, or directly from
% %Eq. \ref{streaming_mass}---\ref{streaming_vz}. 
% Interestingly, we find in dust-rich disks with 
% $\epsilon>1$, Eq. \ref{si_check} can be satisified with $\Delta
% P\simeq \ii\dd v_x \p_rP/\sigma$, i.e. the radial pressure gradient
% is responsible for growth. Conversely, for `linB' with $\epsilon
% < 1$, one can approximate $\Delta P \simeq \dd P$ in Eq. \ref{si_check}.
% This
%suggests that for SI in gas-dominated disks, the distinction between
%Eulerian and Lagragian pressure perturbations is unimportant. 
%{\bf but weak growth in this case}
%However
%in that case the gro

In Table \ref{si_compare} we also calculate the phase difference
between the Lagragian pressure perturbation and density perturbations
as    
\begin{align*} 
\varphi&\equiv \sgn{\left[\real{(\sigma)}\right]}\arg\left(\Delta
  P\Delta\rho^*\right).\notag 
%\notag\\
%       &= \sgn{\left[\real{(\sigma)}\right]}\arg\left(\Delta\rhog^*\Delta\rhod\right).\notag
\end{align*}
(Note that $\Delta P = \delta P + \ii \dd v_x \p_rP/\sigma$, and it is
important to include the global radial pressure gradient here.) Then 
$\varphi > 0 $ indicates gas pressure lagging behind total density, 
which is true for all the cases. Thus SI is indeed associated with
such a phase lag.

@@

In Fig. \ref{si_compare_fig} we calculate the most unstable SI 
mode as a function of $\taus$ at fixed $K_z=30$ and $\epsilon =3$. We
compare results between the full, two-fluid linear analysis and the
one-fluid framework. We also include an analytic model, described in Appendix 
\ref{si_dust_rich}, based on the one-fluid dispersion relation with 
additional approximations (orange diamonds).

Both results based on the one-fluid framework compares well with the 
full, two-fluid analysis, only breaking down at a relatively large
$\taus\OmK\gtrsim 0.1$. For larger $\taus$ the two-fluid phase
lag drops, along with the growth rate. This suggest that a
non-vanishing phase lag is necessary for instability. However, the
magnitude of the phase lag does not correlate with growth rates. In
fact, $\varphi$ remains finite as $|\sigma|, \taus\to0$ (but
non-zero). This arises because the optimum $K_x\propto \taus^{-1/2}$
diverges (see Appendix \ref{si_dust_rich}).

In a future work we will perform a more detailed parameter survey to
compare the simplified one-fluid and full two-fluid frameworks in
calculating SI.

\begin{figure}
\includegraphics[width=\linewidth]{figures/si_2f_1f_compare}
\caption{Comparison of the linear
streaming instability between a full two-fluid analysis (solid
line), the one-fluid framework simplified by the terminal
velocity approximation (green asterisks) and an analytic solution to
the one-fluid dispersion relation in the dust-rich limit (orange
diamonds, see also Appendix \ref{si_dust_rich}). The vertical 
wavenumber is fixed and growth rates are maximized over $K_x$. 
\label{si_compare_fig}}
\end{figure}

%"generalized stokes number" would include other problem parameters,
%e.g. wavenumber

### 2017-05-23 model.tex
wording edits in SI section (main text and new analytic model in appendix)

The two fluids interact via a drag force parameterized by a stopping
time $\tstop$ 
such that 
\begin{align}  
  \rhod\left.\frac{\p \bm{v}_\mathrm{d}}{\p t}\right|_\mathrm{drag}= -
  \rhog\left.\frac{\p \bm{v}_\mathrm{g}}{\p t}\right|_\mathrm{drag}=
  - \frac{\rhog\rhod}{\rho}\frac{\widetilde{\bm{v}}}{\tstop}. 
\end{align}
is the dust-gas friction force per unit volume, where
\begin{align}
  \widetilde{\bm{v}} \equiv\bm{v}_\mathrm{d} - \bm{v}_\mathrm{g} 
\end{align}
is the dust-gas velocity difference; and 
\begin{align}
  \rho \equiv \rhog + \rhod
\end{align}
is the total density. 
Note that $\tstop$ 
differs slightly from the stopping time $\tau_\mathrm{s} =
\tstop\rho/\rhog$ used in some studies
\citep[e.g.][]{youdin05a,jacquet11}.

@@

A parcel of the isothermal dusty-gas mixture does 
postive work if  
\begin{align*}
-\sgn\left(\omega\right)\imag\left(\Delta\rhog\Delta\rhod^*\right)>0,
\end{align*}
meaning that gas follows dust (Fig. \ref{lag_cartoon}). Instabilities
are thus not possible if the gas does not respond to dust 
(i.e. no back-reaction).% {\bf can we say this means that any effective
%  dust-drag instability require high dust to gas ratios?} 
The pressure-density lag shown in
Fig. \ref{lag_cartoon} is achieved if, just after the total density of a 
parcel maximizes, its gas content
is increasing, which requires a sufficiently large
particle flux \emph{out} of the parcel. See $A$ to $B$ in 
Fig. \ref{pdv_cartoon}.

>>>>

The two fluids interact via a drag force parameterized by the relative stopping 
time $\tstop$ such that 
\begin{align}  
  \rhod\left.\frac{\p \bm{v}_\mathrm{d}}{\p t}\right|_\mathrm{drag}= -
  \rhog\left.\frac{\p \bm{v}_\mathrm{g}}{\p t}\right|_\mathrm{drag}=
  - \frac{\rhog\rhod}{\rho}\frac{\widetilde{\bm{v}}}{\tstop}. 
\end{align}
is the dust-gas friction force per unit volume, where
\begin{align}
  \widetilde{\bm{v}} \equiv\bm{v}_\mathrm{d} - \bm{v}_\mathrm{g} 
\end{align}
is the dust-gas velocity difference; and 
\begin{align}
  \rho \equiv \rhog + \rhod
\end{align}
is the total density. 
Note that $\tstop$ 
differs slightly from the \emph{particle} stopping time $\tau_\mathrm{s} =
\tstop\rho/\rhog$ used in some studies
\citep[e.g.][]{youdin05a,jacquet11}.

@@

A parcel of the strictly isothermal dusty-gas mixture does 
postive work if  
\begin{align*}
-\sgn\left(\omega\right)\imag\left(\Delta\rhog\Delta\rhod^*\right)>0,
\end{align*}
meaning that gas follows dust (Fig. \ref{lag_cartoon}). Instabilities
are thus not possible if the gas does not respond to dust 
(i.e. no back-reaction).% {\bf can we say this means that any effective
%  dust-drag instability require high dust to gas ratios?} 
The pressure-density lag shown in
Fig. \ref{lag_cartoon} is achieved if, just after the total density of a 
parcel maximizes, its gas content
is increasing, which requires a sufficiently large
particle flux \emph{out} of the parcel. See $A$ to $B$ in 
Fig. \ref{pdv_cartoon}.
